A Simplified Formulation for Molecular Dynamics with Rigid Bodies

### 2017-02-02 rigid_bodies.tex
canonical ensemble reviewed

In this paper, we report formulation for the motion of rigid bodies in the microcanonical (NVE) and canonical (NVT) ensembles, as well as numerical integrators for both cases. Our contribution in the NVE case resides on a factorization of the rotation matrix that allows a Hamiltonian formalism to be devised without the need of extending the rotational phase space, thus simplifying the formulation of Miller \textit{et al}.\cite{Miller2002} In the NVT case, both the translational and rotational degrees of freedom are simultaneously coupled to a unique Nos\'e-Hoover chain (NHC) thermostat. This differs from the approach of Kamberaj \textit{et al}.,\cite{Kamberaj2005} in which two thermostat chains act separately on the rotational and the translational degrees of freedom. We also simplify the calculation of pressure for systems with rigid bodies whose constituent atoms interact via a pairwise potential under periodic boundary conditions. For this, we derive two alternative expressions: one that depends exclusively on the resultant interaction forces between body pairs and one that corrects the pressure calculated in terms of interactions among freely moving atoms.

The paper is organized as follows. In Sec.~\ref{sec:molecular_dynamics}, we devise numerical integrators for both the NVT dynamics, whose performance and correctness we assess in Sec.~\ref{sec:numerical_results}. Moreover, we analyze the statistical distributions of the sampled energies in the NVT case, and check the equation we developed for computing pressure. Finally, we present some concluding remarks in Sec.~\ref{sec:conclusion}.

\subsection{Microcanonical Ensemble: A Review}
algebraic manipulations
Prompted by the introduction of a particular factorization of the rotational matrix, in a previous paper\cite{Abreu_2017} we reformulated the Hamiltonian equations of motion for rigid bodies, from which we obtained symplectic integrators in the microcanonical (NVE) ensemble.
Nos\'{e}-Hoover chain thermostat. the non-symplectic character, we used the approximate solution for rotation, as we high quality and non-Hamiltonian dynamics admitos no backward error analyisis. We now will briefly review the key expressions of the metioned paper. First, recall that the translational and rotational degrees of freedom are decoupled. Mathematicaly, the position of a particle $j$,  constituent of a rigid body, will vary with time as $\vt r_j(t) = \vt r(t) + \tr{[{\mt A}(\vt q(t))]}\vt d_j$, where $\vt r(t)$ is the position of the center of mass of the body, $\tr{\mt A(\vt q(t))}$ is obtained by transposing the rotational matrix and $\vt d_j$ is the position of the particle in the body-fixed frame of reference. Note that unit quaternions $\vt q = \tr {[\begin{array}{cccc} q_1 & q_2 & q_3 & q_4 \end{array}]}$ are employed to represent the orientations of rigid bodies, which satisfy the known relation\cite{Goldstein2002} $\tr{\vt q}{\vt q} = \|\vt q\|^2 = q_1^2 + q_2^2 + q_3^2 + q_4^2 = 1$, as well as $\tr{\vt q}{\vt q} = 0$.
The matrix $\mt A$ is computed from a given unit quaternion $\vt q$ by\cite{Allen1989,Miller2002}
\begin{equation*}
\label{eq:A_from_q}
\mt A(\vt q) = \begin{smallarray}{ccc}
q_1^2 + q_2^2 - q_3^2 - q_4^2 & 2(q_2 q_3 + q_1 q_4) & 2(q_2 q_4 - q_1 q_3) \\
2(q_2 q_3 - q_1 q_4) & q_1^2 - q_2^2 + q_3^2 - q_4^2 & 2(q_3 q_4 + q_1 q_2) \\
2(q_2 q_4 + q_1 q_3) & 2(q_3 q_4 - q_1 q_2) & q_1^2 - q_2^2 - q_3^2 + q_4^2  
\end{smallarray},
\end{equation*}
which admits the following factorization:
\begin{equation}
\label{eq:factorization_of_A}
{\mt A} = \tr{\mt B}{\mt C}.
\end{equation} 
that depend on the quaternion components as
\begin{equation}
\label{eq:def_B_and_C}
\mt B(\vt q) = \begin{smallarray}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 & -q_4 &  q_3 \\
 q_4 &  q_1 & -q_2 \\
-q_3 &  q_2 &  q_1
\end{smallarray}
\; \text{and} \;
\mt C(\vt q) = \begin{smallarray}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 &  q_4 & -q_3 \\
-q_4 &  q_1 &  q_2 \\
 q_3 & -q_2 &  q_1
\end{smallarray}.
\end{equation}
Note that both $\mt B$ and $\mt C$ depend linearly on the Euler parameters, which makes Eq.~\eqref{eq:factorization_of_A} aconvenient factorization of matrix $\mt A$. This factorization provides a relation between the three-dimensional angular velocity vector $\vt \omega$ and the four-dimensional rates of change of the quaternion components, which is
\begin{equation}
\label{eq:relation_qdot_omega}
\dot{\vt q} = \frac{1}{2} \mt B \vt \omega.
\end{equation}
It avoids the use of an ficticious angular velocity component, as Miller~\textit{et al.}\cite{Miller2002}. An important feature of this relation is that it preserves the unit-norm constraint of a quaternion upon infinitesimal rotation. Note that, for simplicity, we might omit the argument of a matrix-valued function (like $\mt A$, $\vt B$, or $\mt C$) when such argument is $\vt q$.

For a three-dimensional system consisting of $N$ non-collinear rigid bodies that interact via a potential $U(\vt r, \vt q)$, we define the Hamiltonian in terms of its generalized coordinates ($\vt r_i$ and $\vt q_i$) and conjugate momenta ($\vt p_i$ and $\vt \pi_i$), and kinetic energy as follows:

\begin{equation}
\label{eq:H_NVE}
\mathcal{H} = U(\vt r,\vt q) + \sum_{i=1}^N \left(\frac{\tr{\vt p}_i{\vt p}_i}{2m_i} + \frac{\tr{\vt \pi}_i {\mt B}_i {\mt I}_i^{-1} \tr{\mt B}_i {\vt \pi}_i}{8}\right),
\end{equation}

\begin{equation}
\label{eq:H_split_omega}
\mathcal{H} = \frac{\tr{\vt p} \vt p}{2m} + U(\vt r, \vt q) + \frac{1}{2} \sum_{k=1}^3 I_k \omega_k^2.
\end{equation}

where an index $i$ ranging from $1$ to $N$ indicates the quantities related to each body, with $m_i$ is the mass and  $\mt I_i$ the inertia tensor which in the principal frame is a diagonal matrix withe entries given by ($I_1$,$I_2$,$I_3$). The equations of motion 
\begin{subequations}
\label{eq:EDO_system}
\begin{align}
&\dot{\vt r} = \frac{1}{m} \vt p \\
&\dot{\vt p} = \mt F \\
&\dot{\vt q} = \frac{1}{2} \sum_{k=1}^3 \omega_k \hat{\mt B}_k \vt q \label{eq:edo_q} \\
&\dot{\vt \pi} = \frac{1}{2} \sum_{k=1}^3 \omega_k \hat{\mt B}_k \vt \pi + 2 \mt C \vt \tau \label{eq:edo_pi}
\end{align}
\end{subequations}

consider the entries of $\vt \omega$ individually. Given that $2{\mt I}{\vt \omega} = \tr{\mt B}{\vt \pi}$, according to Eq.~\ref{eq:vector_entries}, each entry of $\vt \omega$ is obtained by
\begin{equation}
\label{eq:omega_entry}
\omega_k = \frac{\tr{\vt \pi} \hat{\mt B}_k \vt q}{2 I_k},
\end{equation}
where $\hat{\mt B}_k$ is one of the permutation matrices presented in Appendix~\ref{sec:auxiliary_math} of the aforementione peper.

%Finally, we open the products $\mt B \vt \omega$ and $\mt \Omega \vt \omega$ as in Eq.~\ref{eq:product_B_vector}, which leads us to the following new form for the equations of motion in Eqs.~\ref{eq:EDO_q} and \ref{eq:EDO_pi}:
%\begin{subequations}
%\label{eq:EDO_system_alternative}
%\begin{align}
%&\dot{\vt q} = \frac{1}{2} \sum_{k=1}^3 \omega_k \hat{\mt B}_k \vt q \label{eq:edo_q} \\
%&\dot{\vt \pi} = \frac{1}{2} \sum_{k=1}^3 \omega_k \hat{\mt B}_k \vt \pi + 2 \mt C \vt \tau %\label{eq:edo_pi}
%\end{align}
%\end{subequations}

The phase space of the whole system contains $3N$ spatial coordinates and $4N$ quaternion components, along with their $7N$ conjugate momenta. Of course, $2N$ implicit constraints are known to exist, as $\tr{\vt q}_i{\vt q}_i = 1$ and $\tr{\vt q}_i{\vt \pi}_i = 0$ for all $i$. A Hamiltonian dynamics in phase space will have $\mathcal{H} = E$ as an integral of motion. The dynamics of this system is governed by a Liouville operator with individual operators defined as
\[
\begin{split}
i\!L_b &= \sum_{i=1}^N \left( \tr{\vt F}_i \diff{}{\vt p_i} + 2 \tr{\vt \tau}_i \tr{\mt C}_i \diff{}{\vt \pi_i} \right), \\
i\!L_t &= \sum_{i=1}^N \frac{\tr{\vt p}_i}{m_i}\diff{}{\vt r_i}, \quad \text{and} \\
i\!L_k &= \frac{1}{2} \sum_{i=1}^N {\omega_i}_k \left( \tr{\vt q}_i\tr{\hat{\mt B}_k} \diff{}{\vt q_i} + \tr{\vt \pi_i}\tr{\hat{\mt B}_k} \diff{}{\vt \pi_i} \right).
\end{split}
\]

Therefore, a symplectic integrator can be devised using a splitting scheme with the action of each propagator (boost, free translation, and free uniaxial rotation) occurring simultaneously for all rigid bodies in the system. If there are no external forces acting on the bodies, then additional conserved quantities are the total linear momentum $\sum_{i=1}^N \vt p_i = \vt P$ and the generator of infinitesimal Galilean boosts\cite{Ray1999, Schwichtenberg2015} $\sum_{i=1}^N(\vt p_i t - m_i \vt r_i) = \boldsymbol{\mathcal G}$. If periodic boundary conditions are employed, as it is usual in molecular dynamics simulations, then there is no rotational symmetry and, consequently, no conservation of the total angular momentum.

By defining $i\!L_r = \sum_{k=1}^3 i\!L_k$ for a complete free rotation, the Liouville operator corresponding to the original case of Eq.~\eqref{eq:EDO_system} (as best noticed in its alternative form, Eq.~\eqref{eq:EDO_system_alternative}) can be written as
\begin{equation}
\label{eq:full_operator_uniaxial}
i\!L = i\!L_t + i\!L_b + i\!L_r,
\end{equation}
where $i\!L_t$ corresponds to the first term in the right-hand of Eq.~\eqref{eq:tras_rot_op}. The factorization we employ here for the Liouville operator in Eq.~\eqref{eq:full_operator} is
\begin{equation}
\label{eq:trotter_splitting_NVE}
e^{\Delta t i\!L_{NVE}} = e^{\frac{\Delta t}{2} i\!L_b} e^{\Delta t i\!L_r} e^{\Delta t i\!L_t} e^{\frac{\Delta t}{2} i\!L_b}.
\end{equation}

Update of forces and torques is required only after the free translation and free rotation have been carried out. Following Miller \textit{et al}.,\cite{Miller2002} each complete rotation is split as
\begin{equation}
\label{eq:splitting_rot}
e^{\Delta t i\!L_r} = \prod_{j=1}^n \left( e^{\frac{\delta t}{2} i\!L_3} e^{\frac{\delta t}{2} i\!L_2} e^{\delta t i\!L_1} e^{\frac{\delta t}{2} i\!L_2} e^{\frac{\delta t}{2} i\!L_3} \right),
\end{equation}
where ${\delta t} = {\Delta t}/n$. This means that free rotation, which is computationally inexpensive for not involving force calculations, can be carried out with a smaller time step for accuracy improvement.

@@

We employ the technique of Sergi and Ferrario\cite{Sergi2001} to couple a single Nos\'e-Hoover chain\cite{Martyna1992} (NHC) composed of $M$ thermostats to a system of $N$ rigid bodies. This is simpler than the approach of Kamberaj \textit{et al}.,\cite{Kamberaj2005} who employed two thermostat chains independently coupled to the translational and rotational degrees of freedom. To this end, we consider extra coordinates $\eta_j$ and their conjugate momenta $p_{\eta_j}$, for $j$ ranging from $1$ to $M$. The conserved ``extended Hamiltonian'' in this case is
\begin{equation}
\label{eq:H_nvt}
H = \mathcal{H}(\vt r,\vt p,\vt q,\vt \pi) + \sum_{j=1}^{M}\frac{p_{\eta_j}^2}{2Q_j} + L k_b T\eta_1 + k_b T\sum_{j=2}^M \eta_j,
\end{equation}
where $\mathcal H$ is the same as in Eq.~\ref{eq:H_NVE}, $k_b T$ is the Boltzmann constant multiplied by the system temperature, $L$ is a constant to be determined, and each $Q_j$ is an inertial parameter of thermostat $j$. As recommended in Ref.~\onlinecite{Martyna1992}, one can make $Q_1 = L k_b T t_d^2$ and $Q_j = k_b T t_d^2$ for $j \geq 2$, where $t_d$ is a characteristic time scale of the system. The equations of motion regarding generalized coordinates are identical to those of a Hamiltonian system, i.e. $\dot{\vt r}_i = {\partial H}/{\partial \vt p_i}$ and $\dot{\vt q}_i = {\partial H}/{\partial \vt \pi_i}$ for every particle $i$ and $\dot{\eta}_j = {\partial H}/{\partial {p_\eta}_j}$ for every thermostat $j$. However, the equations of motion for generalized momenta contain non-Hamiltonian terms. For instance, the first thermostat of the chain acts on the translational and rotational momenta of each particle $i$ as
\[
\begin{split}
\dot{\vt p}_i &= -\diff{H}{\vt r_i} - {\vt p}_i \diff{H}{p_{\eta_1}} \quad \text{and} \\
\dot{\vt \pi}_i &= -\diff{H}{\vt q_i} - {\vt \pi}_i \diff{H}{p_{\eta_1}}.
\end{split}
\]

@@

Finally, as the last thermostat will just act on the previous one, its equation of motion can be obtained by simply replacing $j = M$ and removing the last term in the equation above. Therefore, if we take all the required derivatives (including those described in Sec.~\ref{sec:hamiltonian}), the equations of motion for a system of rigid bodies coupled to a single Nos\'e-Hoover chain become
\begin{subequations}
\label{eq:nhc_system}
\begin{align}
&\dot{\vt r}_i = \frac{{\vt p}_i}{m_i} \\ 
&\dot{\vt p}_i = {\vt F}_i - \alpha_1 \vt p_i \label{eq:nhc_p} \\
&\dot{\vt q}_i = \frac{1}{2} \mt B_i \vt \omega_i \label{eq:nhc_q} \\
&\dot{\vt \pi}_i = \frac{1}{2} \mt \Omega_i \vt \omega_i + 2 \mt C_i \vt \tau_i - \alpha_1 \vt \pi_i \label{eq:nhc_pi} \\
&\dot{\eta}_j = \alpha_j \label{eq:nhc_eta} \\
&{\dot p}_{\eta_j} = G_j - \alpha_{j+1} p_{\eta_j}  \label{eq:nhc_p_eta}
\end{align}
\end{subequations}

In the equations above, ${\vt \omega}_i = \frac{1}{2} {\mt I}_i^{-1} \tr{\mt B}_i {\vt \pi}_i$, $\alpha_j = {p_{\eta_j}}/{Q_j}$ for $j \leq M$ and $\alpha_{M+1} = 0$, while
\begin{align*}
&G_1 = \sum_{i=1}^N \left( \frac{\tr{\vt p}_i{\vt p}_i}{m_i} + \tr{\vt \omega}_i \mt I_i \vt \omega_i \right) - L k_b T \quad \text{and}\\
&G_j = \frac{p_{\eta_{j-1}}^2}{Q_{j-1}} - k_b T \qquad \text{for} \; j > 1.
\end{align*}

>>>>

In this paper, we report formulation for the motion of rigid bodies in the microcanonical (NVE) and canonical (NVT) ensembles, as well as numerical integrators for both cases. Our contribution in the NVE case resides on a factorization of the rotation matrix that allows a Hamiltonian formalism to be devised without the need of extending the rotational phase space, thus simplifying the formulation of Miller \textit{et al}.\cite{Miller2002} In the NVT case, both the translational and rotational degrees of freedom are simultaneously coupled to a unique Nos\'e-Hoover chain (NHC) thermostat. This differs from the approach of Kamberaj \textit{et al}.,\cite{Kamberaj2005} in which two thermostat chains act separately on the rotational and the translational degrees of freedom. We also simplify the calculation of pressure for systems with rigid bodies whose constituent atoms interact via a pairwise potential under periodic boundary conditions. For this, we derive two alternative expressions: %one that depends exclusively on the resultant interaction forces between body pairs and one that corrects the pressure calculated in terms of interactions among freely moving atoms.

%The paper is organized as follows. In Sec.~\ref{sec:molecular_dynamics}, we devise numerical integrators for both the NVT dynamics, whose performance and correctness we assess in Sec.~\ref{sec:numerical_results}. Moreover, we analyze the statistical distributions of the sampled energies in the NVT case, and check the equation we developed for computing pressure. Finally, we present some concluding remarks in Sec.~\ref{sec:conclusion}.

\section{Theory}

Prompted by a particular factorization of the rotational matrix, in a previous paper\cite{Abreu_2017} we have reformulated the Hamiltonian equations of motion for rigid bodies, from which we obtained symplectic integrators in the microcanonical (NVE) ensemble. In the aforementioned paper, we introduced an analytical solution for torque-free rotations, which would be particularly useful if any backward error analysis\cite{Hairer_2006} is to be carried out. However, the  Nos\'{e}-Hoover chain thermostat introduces non-Hamiltonian components, so any backward error analysis it is not feasible anymore. Given that, in this paper, the thermostatted version of the equations of motion will be generated considering the approximate solution for the rotational motion, which is also a mathematical simplification of the formulation introduced by Miller~\textit{et al.}\cite{Miller2002}, and exhibits a precision notably similar to that achieved using the analytical solution. In the next subsection, we will skip over the mathematical details and restrict ourselves to some fundamental expressions on which the dynamics in the canonical (NVT) ensemble will be developed. For full details in the NVE ensemble the reader is referred to our previous paper.\cite{Abreu_2017}

\subsection{Microcanonical Ensemble}

When describing the motion of rigid bodies, first recall that the translational and rotational degrees of freedom are decoupled. Mathematically, the position of a particle $j$ constituent of a body, will vary with time as $\vt r_j(t) = \vt r(t) + \tr{[{\mt A}(\vt q(t))]}\vt d_j$, where $\vt r(t)$ is the center of mass position of the body, $\vt d_j$ is the particle position in the body-fixed frame of reference, and $\tr{\mt A(\vt q(t))}$ is obtained by transposing the rotational matrix $\mt A$, which in turn is a function of the unit quaternions $\vt q = \tr {[\begin{array}{cccc} q_1 & q_2 & q_3 & q_4 \end{array}]}$. In terms of $\vt q$, the dynamics must satisfy the known relation\cite{Goldstein2002} $\tr{\vt q}{\vt q} = \|\vt q\|^2 = q_1^2 + q_2^2 + q_3^2 + q_4^2 = 1$, as well as $\tr{\vt q}{\dot{\vt q}} = 0$.

The matrix $\mt A$ is computed from a given unit quaternion $\vt q$ by\cite{Allen1989,Miller2002}
\begin{equation*}
\label{eq:A_from_q}
\mt A(\vt q) = \begin{smallarray}{ccc}
q_1^2 + q_2^2 - q_3^2 - q_4^2 & 2(q_2 q_3 + q_1 q_4) & 2(q_2 q_4 - q_1 q_3) \\
2(q_2 q_3 - q_1 q_4) & q_1^2 - q_2^2 + q_3^2 - q_4^2 & 2(q_3 q_4 + q_1 q_2) \\
2(q_2 q_4 + q_1 q_3) & 2(q_3 q_4 - q_1 q_2) & q_1^2 - q_2^2 - q_3^2 + q_4^2  
\end{smallarray},
\end{equation*}
which admits the following factorization:
\begin{equation}
\label{eq:factorization_of_A}
{\mt A} = \tr{\mt B}{\mt C},
\end{equation} 
where
\begin{equation}
\label{eq:def_B_and_C}
\mt B(\vt q) = \begin{smallarray}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 & -q_4 &  q_3 \\
 q_4 &  q_1 & -q_2 \\
-q_3 &  q_2 &  q_1
\end{smallarray}
\; \text{and} \;
\mt C(\vt q) = \begin{smallarray}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 &  q_4 & -q_3 \\
-q_4 &  q_1 &  q_2 \\
 q_3 & -q_2 &  q_1
\end{smallarray}.
\end{equation}
Note that both $\mt B$ and $\mt C$ depend linearly on the Euler parameters, which makes Eq.~\eqref{eq:factorization_of_A} a convenient factorization of matrix $\mt A$. This factorization provides a relation between the three-dimensional angular velocity vector $\vt \omega$ and the four-dimensional rate of change of the quaternion components, which is
\begin{equation}
\label{eq:relation_qdot_omega}
\dot{\vt q} = \frac{1}{2} \mt B \vt \omega.
\end{equation}
The equation above avoids the introduction of a ficticious angular velocity component, as required in the formulation of Miller~\textit{et al.}\cite{Miller2002} An important feature of this relation is that it preserves the unit-norm constraint of a quaternion upon infinitesimal rotation. For the sake of brevity, the argument of a matrix-valued function (like $\mt A$, $\vt B$, or $\mt C$) might be omitted, when such argument is $\vt q$.

In the Hamiltonian approach, we employ the linear momentum $\vt p = m \dot{\vt r}$, conjugated to $\vt r$, as well as the momentum $\vt \pi= 4 \mt B \mt I \tr{\mt B} \dot{\vt q} = 2 \mt B \mt I \vt \omega$, conjugated to $\vt q$\cite{Goldstein2002}. In this equations, $m$ is the mass of the body and $\mt I$ is the inertia tensor, which is a digonal matrix in the principal frame. For a three-dimensional system consisting of $N$ non-collinear rigid bodies that interact via a potential $U(\vt r, \vt q)$, we define the Hamiltonian in terms of its generalized coordinates ($\vt r_i$ and $\vt q_i$) and conjugate momenta ($\vt p_i$ and $\vt \pi_i$), as follows:~\cite{Abreu_2017}

\begin{equation}
\label{eq:H_NVE}
\mathcal{H} = U(\vt r,\vt q) + \sum_{i=1}^N \left(\frac{\tr{\vt p}_i{\vt p}_i}{2m_i} + \frac{\tr{\vt \pi}_i {\mt B}_i {\mt I}_i^{-1} \tr{\mt B}_i {\vt \pi}_i}{8}\right),
\end{equation}
where the index $i$ ranging from $1$ to $N$ indicates the quantities related to each body.

The equations of motion are obtained from the Hamiltonian by $\dot{\vt r}_i = \partial \mathcal{H} / \partial \vt p_i$, $\dot{\vt p}_i = -\partial \mathcal{H} / \partial \vt r_i$, $\dot{\vt q}_i = \partial \mathcal{H} / \partial \vt \pi_i$, and $\dot{\vt \pi}_i = -\partial \mathcal{H} / \partial \vt q_i$.\cite{Goldstein2002}. For a rigid body composed of $n_p$ individual particles, the derivative $\partial U/\partial \vt r$ is obtained by
\[
\diff{U}{\vt r_i} = -\vt F_i = -\sum_{j=1}^{n_p} {\vt F_j},
\]
where $\vt F_j$ is the force applied on particle $j$, expressed in the space-fixed frame of reference and thus $\vt F_i$ is the resultant force on the body. As demonstrated in Abreu~\textit{et al.},\citep{Abreu_2017} $\partial{U}/\partial \vt q_i$ is given by
\[
\diff{U}{\vt q_i} = -2 \mt C_i \vt \tau_i = -2 \mt C_i \sum_{j=1}^{n_p} {\vt \delta_j} \times {\vt F_j},
\]
where $\vt \tau_i$ is the resultant torque exerted on the body and ${\vt \delta}_j$ is the position of the particle $j$, both expressed in the space-fixed frame.  The position ${\vt \delta}_j$ is related to ${\vt d}_j$ by ${\vt \delta}_j = \tr{\mt A} {\vt d}_j$.\cite{Goldstein2002} After carrying out the required differentiations of $\mathcal{H}$  we obtain the following Hamiltonian system of equations of motion for a rigid body:
\begin{subequations}
\label{eq:EDO_system}
\begin{align}
&\dot{\vt r}_i = \frac{1}{m_i} \vt p_i \\
&\dot{\vt p}_i = \mt F_i \\
&\dot{\vt q}_i = \frac{1}{2} \mt B_i \vt \omega_i \label{eq:EDO_q} \\
&\dot{\vt \pi}_i = \frac{1}{2} \mt \Omega_i \vt \omega_i + 2 \mt C_i \vt \tau_i \label{eq:EDO_pi}
\end{align}
\end{subequations}
with $\mt \Omega_i = \mt B_i(\vt \pi_i)$.

The approximate solution for a complete rotation is obtained by opening the products $\mt B \vt \omega$ and $\mt \Omega \vt \omega$ employing each entry of $\vt \omega$: 
\begin{equation}
\label{eq:omega_entry}
\omega_k = \frac{\tr{\vt \pi} \hat{\mt B}_k \vt q}{2 I_k},
\end{equation}
where $\hat{\mt B}_k$ is one of the following permutation matrices:
\[
\hat{\mt B}_1 = \begin{smallarray}{cccc}
 0 & \text{-}1 &  0 &  0 \\
 1 &  0 &  0 &  0 \\
 0 &  0 &  0 &  1 \\
 0 &  0 & \text{-}1 &  0 \\
\end{smallarray} \quad
\hat{\mt B}_2 = \begin{smallarray}{cccc}
 0 &  0 & \text{-}1 &  0 \\
 0 &  0 &  0 & \text{-}1 \\
 1 &  0 &  0 &  0 \\
 0 &  1 &  0 &  0 \\
\end{smallarray} \quad
\hat{\mt B}_3 = \begin{smallarray}{cccc}
 0 &  0 &  0 & \text{-}1 \\
 0 &  0 &  1 &  0 \\
 0 & \text{-}1 &  0 &  0 \\
 1 &  0 &  0 &  0 \\
\end{smallarray}
\]
The new form of the equations of motion in Eqs.~\eqref{eq:EDO_q} and \eqref{eq:EDO_pi} are then:
\begin{subequations}
\label{eq:EDO_system_alternative}
\begin{align}
&\dot{\vt q}_i = \frac{1}{2} \left( \sum_{k=1}^3 \omega_{ik} \hat{\mt B}_k \right) \vt q_i \label{eq:edo_q} \\
&\dot{\vt \pi}_i = \frac{1}{2} \left( \sum_{k=1}^3 \omega_{ik} \hat{\mt B}_k \right) \vt \pi_i + 2 \mt C_i \vt \tau_i \label{eq:edo_pi}
\end{align}
\end{subequations}

In this context, the Hamiltonian of the rigid body can be rewritten as
\begin{equation}
\label{eq:H_split_omega}
\mathcal{H} =  U(\vt r, \vt q) +  \sum_{i=1}^N \left( \frac{\tr{\vt p_i} \vt p_i}{2m_i} + \frac{1}{2} \sum_{k=1}^3 I_{ik} \omega_{ik}^2\right),
\end{equation}
from which we recognize partial Hamiltonians, each of one generates equations of motion with known analytical solutions. In particular, we have $\mathcal{H}_b = U(\vt r, \vt q)$, $\mathcal{H}_{t} = \frac{1}{2m}\tr{\vt p} \vt p$, and $\mathcal{H}_k = (1/2) I_k \omega_k^2$. Of coarse, $\mathcal{H} = \mathcal{H}_b + \mathcal{H}_t + \sum_{k=1}^3 \mathcal{H}_k$. The NVE dynamics of this system is governed by a Liouville operator given by\cite{Abreu_2017}
\begin{equation}
\label{eq:full_operator}
i\!L_{NVE} = i\!L_t + i\!L_b + i\!L_r,
\end{equation}
where the individual operators are defined as follows:
\[
\begin{split}
&boost:\\
i\!L_b &= \sum_{i=1}^N \left( \tr{\vt F}_i \diff{}{\vt p_i} + 2 \tr{\vt \tau}_i \tr{\mt C}_i \diff{}{\vt \pi_i} \right), \\
&translation: \\
 i\!L_t &= \sum_{i=1}^N \frac{\tr{\vt p}_i}{m_i}\diff{}{\vt r_i}, \quad \text{and} \\
&uniaxial \; rotation:\\
 i\!L_k &= \frac{1}{2} \sum_{i=1}^N {\omega_i}_k \left( \tr{\vt q}_i\tr{\hat{\mt B}_k} \diff{}{\vt q_i} + \tr{\vt \pi_i}\tr{\hat{\mt B}_k} \diff{}{\vt \pi_i} \right),
\end{split}
\]
using $i\!L_r = \sum_{k=1}^3 i\!L_k$ for a complete torque-free rotation.
The splitting of the Liouville operator of Eq.~\eqref{eq:full_operator} we employ in this paper is
\begin{equation}
\label{eq:trotter_splitting_NVE}
e^{\Delta t i\!L_{NVE}} = e^{\frac{\Delta t}{2} i\!L_b} e^{\Delta t i\!L_r} e^{\Delta t i\!L_t} e^{\frac{\Delta t}{2} i\!L_b},
\end{equation}
Update of forces and torques is required only after the free translation and free rotation have been carried out. Following Miller \textit{et al}.,\cite{Miller2002} each complete rotation is splitted as
\begin{equation}
\label{eq:splitting_rot}
e^{\Delta t i\!L_r} =  e^{\frac{\Delta t}{2} i\!L_3} e^{\frac{\Delta t}{2} i\!L_2} e^{\Delta t i\!L_1} e^{\frac{\Delta t}{2} i\!L_2} e^{\frac{\Delta t}{2} i\!L_3},
\end{equation}
where ${\Delta t}$ is the time step size.

As demonstrated in our previous paper,\cite{Abreu2017}, the action of the propagators 
\begin{align*}
&{\vt r} = {\vt r}_0 + (t/m) {\vt p}_0,
\end{align*}

\begin{align*}
&{\vt p} = {\vt p}_0 + {\vt F}_0 t, \\
&{\vt \pi} = {\vt \pi}_0 +  2 {\mt C}_0 {\vt \tau}_0 t.
\end{align*}

\begin{align*}
&{\vt q}(t) = \cos\left(\frac{\omega_k t}{2}\right) \vt q_0 + \sin\left(\frac{\omega_k t}{2}\right) \hat{\mt B}_k \vt q_0 \ \text{and} \\
&{\vt \pi}(t) = \cos\left(\frac{\omega_k t}{2}\right) \vt \pi_0 + \sin\left(\frac{\omega_k t}{2}\right) \hat{\mt B}_k \vt \pi_0, 
\end{align*}

@@

We employ the technique of Sergi and Ferrario\cite{Sergi2001} to couple a single Nos\'e-Hoover chain\cite{Martyna1992} (NHC) composed of $M$ thermostats to a system of $N$ rigid bodies. This is simpler than the approach of Kamberaj \textit{et al}.,\cite{Kamberaj2005} who employed two thermostat chains independently coupled to the translational and rotational degrees of freedom. To this end, we consider extra coordinates $\eta_j$ and their conjugate momenta $p_{\eta_j}$, for $j$ ranging from $1$ to $M$. The conserved ``extended Hamiltonian'' in this case is
\begin{equation}
\label{eq:H_nvt}
H = \mathcal{H}(\vt r,\vt p,\vt q,\vt \pi) + \sum_{j=1}^{M}\frac{p_{\eta_j}^2}{2Q_j} + L k_b T\eta_1 + k_b T\sum_{j=2}^M \eta_j,
\end{equation}
where $\mathcal H$ is the same as in Eq.~\eqref{eq:H_NVE}, $k_b T$ is the Boltzmann constant multiplied by the system temperature, $L$ is a constant to be determined, and each $Q_j$ is an inertial parameter of thermostat $j$. As recommended in Ref.~\onlinecite{Martyna1992}, one can make $Q_1 = L k_b T t_d^2$ and $Q_j = k_b T t_d^2$ for $j \geq 2$, where $t_d$ is a characteristic time scale of the system. The equations of motion regarding generalized coordinates are identical to those of a Hamiltonian system, i.e. $\dot{\vt r}_i = {\partial H}/{\partial \vt p_i}$ and $\dot{\vt q}_i = {\partial H}/{\partial \vt \pi_i}$ for every particle $i$ and $\dot{\eta}_j = {\partial H}/{\partial {p_\eta}_j}$ for every thermostat $j$. However, the equations of motion for generalized momenta contain non-Hamiltonian terms. For instance, the first thermostat of the chain acts on the translational and rotational momenta of each particle $i$ as
\[
\begin{split}
\dot{\vt p}_i &= -\diff{H}{\vt r_i} - {\vt p}_i \diff{H}{p_{\eta_1}} \quad \text{and} \\
\dot{\vt \pi}_i &= -\diff{H}{\vt q_i} - {\vt \pi}_i \diff{H}{p_{\eta_1}}.
\end{split}
\]

@@

Given that the last thermostat will just act on the previous one, its equation of motion can be obtained by simply replacing $j = M$ and removing the last term in the equation above. Therefore, the equations of motion for a system of rigid bodies coupled to a single Nos\'e-Hoover chain become
\begin{subequations}
\label{eq:nhc_system}
\begin{align}
&\dot{\vt r}_i = \frac{{\vt p}_i}{m_i} \\ 
&\dot{\vt p}_i = {\vt F}_i - \alpha_1 \vt p_i \label{eq:nhc_p} \\
&\dot{\vt q}_i = \frac{1}{2} \mt B_i \vt \omega_i \label{eq:nhc_q} \\
&\dot{\vt \pi}_i = \frac{1}{2} \mt \Omega_i \vt \omega_i + 2 \mt C_i \vt \tau_i - \alpha_1 \vt \pi_i \label{eq:nhc_pi} \\
&\dot{\eta}_j = \alpha_j \label{eq:nhc_eta} \\
&{\dot p}_{\eta_j} = G_j - \alpha_{j+1} p_{\eta_j}  \label{eq:nhc_p_eta}
\end{align}
\end{subequations}

In the equations above, ${\vt \omega}_i$ $=$ $\frac{1}{2} {\mt I}_i^{-1} \tr{\mt B}_i {\vt \pi}_i$, $\alpha_j$ $=$ ${p_{\eta_j}}/{Q_j}$ for $j \leq M$ and $\alpha_{M+1} = 0$, while
\begin{align*}
&G_1 = \sum_{i=1}^N \left( \frac{\tr{\vt p}_i{\vt p}_i}{m_i} + \tr{\vt \omega}_i \mt I_i \vt \omega_i \right) - L k_b T \quad \text{and}\\
&G_j = \frac{p_{\eta_{j-1}}^2}{Q_{j-1}} - k_b T \qquad \text{for} \; j > 1.
\end{align*}

Note that the $3N$ momentum components in Eq.~\eqref{eq:nhc_p}, the $4N$ components in Eq.~\eqref{eq:nhc_pi}, and the thermostat momenta in Eq.~\eqref{eq:nhc_p_eta} affect their own time-derivatives. Thus, the thermostatted system has a phase-space compressibility $\kappa = -7N \alpha_1 - \sum_j \alpha_{j+1}$. Resorting to Eq.~\eqref{eq:nhc_eta} and the fact that $\alpha_{M+1} = 0$, the function $w(\vt x)$ that determines the invariant phase-space measure is given by
\begin{equation}
\label{eq:nhc_measure}
w = -7N \eta_1 - \sum_{j=2}^M \eta_j.
\end{equation}
